---
title: "Cloud-Native Observability"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cloud-Native Observability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Integrate securetrace with OTLP collectors, Prometheus, and distributed tracing
via W3C Trace Context. For core tracing, see `vignette("observability")`.

## OTLP Export

Inspect the OTLP payload locally with `otlp_format_trace()`.

```{r}
library(securetrace)

tr <- Trace$new("format-demo")
tr$start()
s <- Span$new("llm-call", type = "llm")
s$start()
s$set_model("claude-sonnet-4-5")
s$set_tokens(input = 1500L, output = 300L)
s$end()
tr$add_span(s)
tr$end()

otlp <- otlp_format_trace(tr$to_list(), service_name = "my-agent")
str(otlp, max.level = 3)
```

Point `exporter_otlp()` at any OTLP-HTTP endpoint.

```{r, eval = FALSE}
exp <- exporter_otlp(endpoint = "http://localhost:4318")
with_trace("data-pipeline", exporter = exp, {
  with_span("fetch", type = "tool", { record_latency(0.25) })
  with_span("analyze", type = "llm", {
    record_tokens(2000, 500, model = "claude-sonnet-4-5")
  })
})
```

Pass headers for authenticated collectors.

```{r}
exp <- exporter_otlp(
  endpoint = "https://tempo.example.com:4318",
  headers = list(Authorization = "Bearer <token>"),
  service_name = "my-r-agent"
)
```

Buffer traces with `batch_size`; call `flush_otlp()` at exit to drain.

```{r, eval = FALSE}
exp <- exporter_otlp("http://localhost:4318", batch_size = 10, max_retries = 3)
for (i in seq_len(5)) {
  with_trace(paste0("run-", i), exporter = exp, {
    with_span("work", type = "tool", { i })
  })
}
flush_otlp(exp)
```

## Prometheus

Create a registry, feed it a trace, render text format.

```{r}
reg <- prometheus_registry()
tr <- Trace$new("agent-run")
tr$start()
s1 <- Span$new("planning", type = "llm")
s1$start()
s1$set_model("claude-sonnet-4-5")
s1$set_tokens(input = 3000L, output = 800L)
s1$end()
tr$add_span(s1)
s2 <- Span$new("execute", type = "tool")
s2$start()
s2$end()
tr$add_span(s2)
tr$end()
prometheus_metrics(tr, reg)
cat(format_prometheus(reg))
```

Use `exporter_prometheus()` so completed traces auto-populate the registry.

```{r}
reg <- prometheus_registry()
prom_exp <- exporter_prometheus(reg)
with_trace("run-1", exporter = prom_exp, {
  with_span("llm", type = "llm", {
    record_tokens(1000, 200, model = "claude-haiku-4-5")
    "done"
  })
})
with_trace("run-2", exporter = prom_exp, {
  with_span("llm", type = "llm", {
    record_tokens(2000, 400, model = "claude-sonnet-4-5")
    "done"
  })
})
cat(format_prometheus(reg))
```

Serve `/metrics` for Prometheus to scrape.

```{r, eval = FALSE}
srv <- serve_prometheus(reg, host = "0.0.0.0", port = 9090)
# ... run traced workloads ...
httpuv::stopServer(srv)
```

## W3C Trace Context

Generate and parse W3C `traceparent` headers.

```{r}
tp <- traceparent(
  trace_id = "4bf92f3577b34da6a3ce929d0e0e4736",
  span_id = "00f067aa0ba902b7"
)
tp
traceparent("4bf92f3577b34da6a3ce929d0e0e4736", "00f067aa0ba902b7", sampled = FALSE)
```

```{r}
ctx <- parse_traceparent("00-4bf92f3577b34da6a3ce929d0e0e4736-00f067aa0ba902b7-01")
ctx$trace_id
ctx$span_id
ctx$sampled
parse_traceparent("invalid-header")
```

Inject `traceparent` into outgoing requests from inside an active trace.

```{r}
with_trace("http-client", {
  with_span("api-call", type = "tool", {
    headers <- inject_headers(list("Content-Type" = "application/json"))
    headers$traceparent
  })
})
```

Extract trace context from incoming headers (case-insensitive).

```{r}
incoming <- list(
  "Content-Type" = "application/json",
  traceparent = "00-4bf92f3577b34da6a3ce929d0e0e4736-00f067aa0ba902b7-01"
)
ctx <- extract_trace_context(incoming)
ctx$trace_id
ctx$span_id
```

## Distributed Tracing with Plumber

Server: extract parent context and create child spans.

```{r, eval = FALSE}
library(securetrace)
library(plumber)

#* @post /analyze
function(req, res) {
  ctx <- extract_trace_context(list(traceparent = req$HTTP_TRACEPARENT))
  tr <- Trace$new("plumber-analyze")
  tr$start()
  s <- Span$new("analyze", type = "llm", parent_id = ctx$span_id)
  s$start()
  s$set_model("claude-sonnet-4-5")
  s$set_tokens(input = 2000L, output = 500L)
  result <- list(status = "ok", answer = 42)
  s$end()
  tr$add_span(s)
  tr$end()
  export_trace(exporter_otlp("http://localhost:4318"), tr)
  res$setHeader("traceparent", traceparent(tr$trace_id, s$span_id))
  result
}
```

Client: inject context so both sides share one trace.

```{r, eval = FALSE}
with_trace("client-workflow", {
  with_span("call-plumber", type = "tool", {
    resp <- httr2::request("http://localhost:8000/analyze") |>
      httr2::req_headers(!!!inject_headers()) |>
      httr2::req_method("POST") |>
      httr2::req_perform()
    httr2::resp_body_json(resp)
  })
})
```
