---
title: "Cloud-Native Observability"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cloud-Native Observability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

securetrace ships with cloud-native observability features that integrate R
agent workflows into modern infrastructure stacks. This vignette covers OTLP
export (Jaeger, Grafana Tempo), Prometheus metrics, and W3C Trace Context
propagation for distributed tracing.

For core tracing concepts (traces, spans, events, exporters), see
`vignette("observability")`.

## How It All Fits Together

In a production deployment, traces flow from your R agent through several
systems. The following diagram shows the typical data path:

```
  R Agent Process                    Infrastructure
  ================                   ==============

  with_trace("run", {
    with_span("llm", ...)   --.
    with_span("tool", ...)  --+--> OTLP exporter
  })                           |       |
                               |       v
                               |   Jaeger / Grafana Tempo
                               |   (distributed trace viewer)
                               |
                               +--> Prometheus exporter
                               |       |
                               |       v
                               |   Prometheus server
                               |   (scrapes /metrics endpoint)
                               |       |
                               |       v
                               |   Grafana dashboards
                               |   (token usage, cost, latency)
                               |
                               +--> JSONL exporter
                                       |
                                       v
                                   Local file
                                   (audit trail, post-hoc analysis)
```

Each export path serves a different purpose: OTLP gives you trace visualization
with span hierarchy, Prometheus gives you time-series metrics for alerting, and
JSONL gives you a durable local record for compliance or batch analysis.

## OTLP Export

The OpenTelemetry Protocol (OTLP) is the standard wire format for sending
traces to collectors like Jaeger, Grafana Tempo, and the OpenTelemetry
Collector. securetrace provides two functions for OTLP integration.

### Sending Traces to a Collector

`otlp_exporter()` creates an exporter that converts traces to OTLP JSON and
POSTs them to a collector endpoint:

```{r}
library(securetrace)

# Export to a local Jaeger instance (default OTLP HTTP port)
exp <- otlp_exporter(endpoint = "http://localhost:4318")

with_trace("data-pipeline", exporter = exp, {
  with_span("fetch", type = "tool", {
    record_latency(0.25)
    data.frame(x = 1:100)
  })

  with_span("analyze", type = "llm", {
    record_tokens(2000, 500, model = "claude-sonnet-4-5")
    "analysis complete"
  })
})
#> Trace exported to http://localhost:4318/v1/traces
```

For authenticated collectors, pass headers:

```{r}
exp <- otlp_exporter(
  endpoint = "https://tempo.example.com:4318",
  headers = list(Authorization = "Bearer <token>"),
  service_name = "my-r-agent"
)
```

The `service_name` parameter sets the `service.name` resource attribute in the
OTLP payload, which is how Jaeger and Tempo group traces by service.

### Inspecting OTLP Output

Before pointing at a live collector, you may want to inspect what securetrace
actually sends. `otlp_format_trace()` is the pure function underneath the
exporter. It converts a trace list into the OTLP `ExportTraceServiceRequest`
JSON structure without sending anything:

```{r}
tr <- Trace$new("format-demo")
tr$start()

s <- Span$new("llm-call", type = "llm")
s$start()
s$set_model("claude-sonnet-4-5")
s$set_tokens(input = 1500L, output = 300L)
s$end()
tr$add_span(s)

tr$end()

otlp <- otlp_format_trace(tr$to_list(), service_name = "my-agent")
str(otlp, max.level = 3)
#> List of 1
#>  $ resourceSpans:List of 1
#>   ..$ :List of 2
#>   .. ..$ resource  :List of 1
#>   .. ..$ scopeSpans:List of 1
```

This is useful for debugging OTLP payloads or writing them to files for
offline import.

### OTLP Batching and Retry

Sending an HTTP request for every single trace is expensive, especially in
high-throughput agent workflows where you might complete dozens of traces per
minute. The OTLP exporter addresses this with two production-essential features:
batching and retry.

**Batching** reduces network overhead by buffering traces in memory and sending
them to the collector in groups. By default, traces are buffered until the
buffer reaches `batch_size`, at which point a single HTTP POST delivers the
entire batch. This means 10 traces cost one HTTP round-trip instead of 10.

**Retry with exponential backoff** handles the reality that collectors
sometimes return transient errors -- rate limits (429), server errors (5xx),
or momentary network failures. Instead of silently dropping traces, the
exporter retries with increasing delays (1s, 2s, 4s, ...) up to `max_retries`
attempts. This makes trace delivery robust without requiring you to build
retry logic yourself.

Use `flush_otlp()` to force-send any buffered traces -- for example, at
process exit to ensure nothing is lost:

```{r}
exp <- otlp_exporter(
  endpoint = "http://localhost:4318",
  batch_size = 10,
  max_retries = 3
)

# Traces accumulate in a buffer...
for (i in seq_len(5)) {
  with_trace(paste0("run-", i), exporter = exp, {
    with_span("work", type = "tool", { i })
  })
}

# Force-send remaining buffered traces before exiting
flush_otlp(exp)
```

A good practice is to call `flush_otlp()` in an `on.exit()` handler or a
`tryCatch(finally = ...)` block so that traces are delivered even if your
script terminates unexpectedly.

## Prometheus Metrics

Prometheus is a pull-based monitoring system. While OTLP sends full trace
detail to a collector, Prometheus works differently: your application exposes
an HTTP endpoint with aggregated metrics, and Prometheus scrapes it at a
regular interval. This makes Prometheus ideal for dashboards and alerting --
"how many tokens did we use in the last hour?" or "alert if error rate exceeds
5%."

securetrace bridges the gap by extracting metrics from traces and exposing
them in Prometheus text exposition format.

### Creating a Registry

A registry holds counters and histograms. Create one and feed traces into it:

```{r}
reg <- prometheus_registry()

# Build a trace
tr <- Trace$new("agent-run")
tr$start()

s1 <- Span$new("planning", type = "llm")
s1$start()
s1$set_model("claude-sonnet-4-5")
s1$set_tokens(input = 3000L, output = 800L)
s1$end()
tr$add_span(s1)

s2 <- Span$new("execute", type = "tool")
s2$start()
s2$end()
tr$add_span(s2)

tr$end()

# Extract metrics from the trace into the registry
prometheus_metrics(tr, reg)
```

### Viewing Metrics

`format_prometheus()` renders the registry as a Prometheus-compatible text
string:

```{r}
cat(format_prometheus(reg))
#> # HELP securetrace_spans_total Total spans by type and status
#> # TYPE securetrace_spans_total counter
#> securetrace_spans_total{type="llm",status="completed"} 1
#> securetrace_spans_total{type="tool",status="completed"} 1
#> # HELP securetrace_tokens_total Total tokens by direction and model
#> # TYPE securetrace_tokens_total counter
#> securetrace_tokens_total{direction="input",model="claude-sonnet-4-5"} 3000
#> securetrace_tokens_total{direction="output",model="claude-sonnet-4-5"} 800
#> # HELP securetrace_traces_total Total traces by status
#> # TYPE securetrace_traces_total counter
#> securetrace_traces_total{status="completed"} 1
```

The registry is cumulative -- each call to `prometheus_metrics()` adds to the
existing counters and histograms. This lets you track metrics across multiple
agent runs.

### Using the Prometheus Exporter

`prometheus_exporter()` returns a securetrace exporter that automatically
feeds each completed trace into a registry:

```{r}
reg <- prometheus_registry()
prom_exp <- prometheus_exporter(reg)

# Traces auto-populate the registry on completion
with_trace("run-1", exporter = prom_exp, {
  with_span("llm", type = "llm", {
    record_tokens(1000, 200, model = "claude-haiku-4-5")
    "done"
  })
})

with_trace("run-2", exporter = prom_exp, {
  with_span("llm", type = "llm", {
    record_tokens(2000, 400, model = "claude-sonnet-4-5")
    "done"
  })
})

# Registry now has cumulative metrics from both runs
cat(format_prometheus(reg))
```

### Serving a /metrics Endpoint

`serve_prometheus()` starts an httpuv HTTP server that Prometheus can scrape:

```{r}
reg <- prometheus_registry()
prom_exp <- prometheus_exporter(reg)

# Start the metrics server
srv <- serve_prometheus(reg, host = "0.0.0.0", port = 9090)

# Run your agent -- metrics accumulate automatically
with_trace("production-run", exporter = prom_exp, {
  with_span("work", type = "llm", {
    record_tokens(5000, 1000, model = "claude-opus-4-6")
    "result"
  })
})

# Prometheus scrapes http://localhost:9090/metrics

# When done, stop the server
httpuv::stopServer(srv)
```

Add the scrape target to your `prometheus.yml`:

```yaml
scrape_configs:
  - job_name: "r-agent"
    static_configs:
      - targets: ["localhost:9090"]
```

## W3C Trace Context Propagation

When R agents call external services (or are called by them), trace context
must propagate across process boundaries so that the full distributed trace
can be reconstructed. securetrace implements the
[W3C Trace Context](https://www.w3.org/TR/trace-context/) standard for this.

### Generating Traceparent Headers

`traceparent()` builds a W3C-compliant header string from trace and span IDs:

```{r}
tp <- traceparent(
  trace_id = "4bf92f3577b34da6a3ce929d0e0e4736",
  span_id = "00f067aa0ba902b7"
)
tp
#> [1] "00-4bf92f3577b34da6a3ce929d0e0e4736-00f067aa0ba902b7-01"
```

The `sampled` parameter controls the trace flags:

```{r}
traceparent(
  "4bf92f3577b34da6a3ce929d0e0e4736",
  "00f067aa0ba902b7",
  sampled = FALSE
)
#> [1] "00-4bf92f3577b34da6a3ce929d0e0e4736-00f067aa0ba902b7-00"
```

### Parsing Incoming Headers

`parse_traceparent()` extracts the components from a traceparent string:

```{r}
ctx <- parse_traceparent(
  "00-4bf92f3577b34da6a3ce929d0e0e4736-00f067aa0ba902b7-01"
)
ctx$trace_id
#> [1] "4bf92f3577b34da6a3ce929d0e0e4736"
ctx$span_id
#> [1] "00f067aa0ba902b7"
ctx$sampled
#> [1] TRUE

# Invalid headers return NULL
parse_traceparent("invalid-header")
#> NULL
```

### Injecting Context into Outgoing Requests

Inside an active trace, `inject_headers()` adds the `traceparent` header
automatically from the current trace and span context:

```{r}
with_trace("http-client", {
  with_span("api-call", type = "tool", {
    headers <- inject_headers(list("Content-Type" = "application/json"))
    headers$traceparent
    #> [1] "00-<trace_id>-<span_id>-01"

    # Use these headers in your HTTP request
    # httr2::req_headers(req, !!!headers)
  })
})
```

### Extracting Context from Incoming Requests

`extract_trace_context()` finds and parses the traceparent header from a
named list (case-insensitive lookup):

```{r}
# Simulate incoming HTTP headers
incoming <- list(
  "Content-Type" = "application/json",
  traceparent = "00-4bf92f3577b34da6a3ce929d0e0e4736-00f067aa0ba902b7-01"
)

ctx <- extract_trace_context(incoming)
ctx$trace_id
#> [1] "4bf92f3577b34da6a3ce929d0e0e4736"
ctx$span_id
#> [1] "00f067aa0ba902b7"
```

## Distributed Tracing with Plumber

Distributed tracing becomes concrete when two services -- a client and a
server -- share the same trace ID. The client injects context into its
request, the server extracts it and creates child spans, and the resulting
trace shows the complete journey across both processes.

Here is a complete example of a Plumber API endpoint that extracts incoming
trace context and creates a child trace for its work:

```{r}
library(securetrace)
library(plumber)

#* @post /analyze
function(req, res) {
  # Extract parent trace context from incoming request
  headers <- list(traceparent = req$HTTP_TRACEPARENT)
  ctx <- extract_trace_context(headers)

  # Create a trace that continues the parent context
  tr <- Trace$new("plumber-analyze")
  tr$start()

  # Create a span linked to the parent
  s <- Span$new("analyze", type = "llm",
                parent_id = ctx$span_id)
  s$start()
  s$set_model("claude-sonnet-4-5")
  s$set_tokens(input = 2000L, output = 500L)

  # Do work...
  result <- list(status = "ok", answer = 42)

  s$end()
  tr$add_span(s)
  tr$end()

  # Export to the same collector as the caller
  exp <- otlp_exporter("http://localhost:4318")
  export_trace(exp, tr)

  # Inject context into response headers for downstream
  res$setHeader(
    "traceparent",
    traceparent(tr$trace_id, s$span_id)
  )

  result
}
```

On the client side, inject context into your requests:

```{r}
with_trace("client-workflow", {
  with_span("call-plumber", type = "tool", {
    headers <- inject_headers()

    resp <- httr2::request("http://localhost:8000/analyze") |>
      httr2::req_headers(!!!headers) |>
      httr2::req_method("POST") |>
      httr2::req_perform()

    httr2::resp_body_json(resp)
  })
})
```

Both the client and server traces share the same `trace_id`, so they appear
as a single distributed trace in Jaeger or Tempo.

## Multi-Exporter Setup

In production, you typically want traces going to multiple destinations. Use
`multi_exporter()` to combine OTLP, Prometheus, and JSONL export:

```{r}
# Set up all three exporters
reg <- prometheus_registry()

combined <- multi_exporter(
  otlp_exporter("http://localhost:4318", service_name = "r-agent"),
  prometheus_exporter(reg),
  jsonl_exporter("traces.jsonl")
)

# Start Prometheus scrape endpoint
srv <- serve_prometheus(reg, port = 9090)

# All traces go to Jaeger + Prometheus + local file
set_default_exporter(combined)

with_trace("production-workflow", {
  with_span("planning", type = "llm", {
    record_tokens(3000, 800, model = "claude-sonnet-4-5")
    "plan ready"
  })

  with_span("execution", type = "tool", {
    record_latency(1.2)
    42
  })
})

# Clean up
httpuv::stopServer(srv)
```

This gives you:

- **Jaeger/Tempo** -- full distributed traces with span hierarchy
- **Prometheus** -- time-series metrics for dashboards and alerting
- **JSONL** -- local audit trail for compliance and post-hoc analysis
