---
title: "Exporting and Analyzing Traces"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exporting and Analyzing Traces}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Getting Traces Out of R

A trace is only valuable if it can be analyzed. During development you might
print traces to the console to see what an agent did. In production you want
traces written to files for post-hoc analysis or piped to an observability
platform. securetrace separates trace *collection* from trace *export* so that
the same instrumentation code works everywhere -- only the exporter changes.

The export pipeline looks like this:

```
with_trace("run", exporter = exp, {
  ...your agent code...
})
    |
    v
Trace$to_list()            # Serialize R6 object to plain list
    |
    v
exporter@export(trace_list) # Deliver to destination
    |
    +---> jsonl_exporter    --> append JSON line to file
    +---> console_exporter  --> print summary to console
    +---> multi_exporter    --> fan out to N exporters
    +---> new_exporter(fn)  --> your custom logic
```

Every exporter receives the same serialized trace list, so they are fully
composable. You can wrap any combination of exporters in a `multi_exporter()`
and every trace goes to every destination.

## JSONL Export

The `jsonl_exporter()` writes each completed trace as a single JSON line to
a file. This is the workhorse exporter for production use -- it produces a
compact, append-only log that is easy to parse with `jsonlite`, `jq`, or any
data pipeline tool.

```{r}
library(securetrace)

# Create a temporary file for this example
trace_file <- tempfile(fileext = ".jsonl")

exp <- jsonl_exporter(trace_file)

# Traces are exported automatically when with_trace() completes
with_trace("exported-run", exporter = exp, {
  with_span("llm-call", type = "llm", {
    record_tokens(1500, 300, model = "claude-sonnet-4-5")
    "response"
  })

  with_span("tool-call", type = "tool", {
    42
  })
})

# Read back the exported data
lines <- readLines(trace_file)
length(lines)  # 1 line per trace

# Parse and inspect
trace_data <- jsonlite::fromJSON(lines[[1]])
trace_data$name
trace_data$status
length(trace_data$spans$span_id)

# Clean up
unlink(trace_file)
```

Each JSONL line contains the full trace structure: trace ID, name, status,
timestamps, duration, and all spans with their tokens, events, and metrics.

## Console Exporter

The `console_exporter()` prints trace summaries directly to the console.
This is the fastest way to see what an agent did during interactive
development, without writing any files.

```{r}
debug_exp <- console_exporter(verbose = TRUE)

with_trace("debug-run", exporter = debug_exp, {
  with_span("planning", type = "llm", {
    record_tokens(2000, 500, model = "claude-sonnet-4-5")
    "plan"
  })

  with_span("execution", type = "tool", {
    100
  })
})
```

Set `verbose = FALSE` to print only the trace header without individual span
details. This is useful in loops or batch runs where you want confirmation that
traces completed without flooding the console.

## Multi-Exporter Setup

In practice you usually want traces going to more than one place -- a file for
the audit trail, the console for visibility, and perhaps a cloud exporter for
production monitoring. `multi_exporter()` fans out to any number of exporters:

```{r}
trace_file <- tempfile(fileext = ".jsonl")

# Combine a file exporter and a console exporter
file_exp <- jsonl_exporter(trace_file)
console_exp <- console_exporter(verbose = TRUE)
combined <- multi_exporter(file_exp, console_exp)

with_trace("multi-export-run", exporter = combined, {
  with_span("llm-call", type = "llm", {
    record_tokens(3000, 600, model = "claude-opus-4-6")
    "result"
  })
})

# The trace was written to the file AND printed to console
lines <- readLines(trace_file)
length(lines)

unlink(trace_file)
```

## Default Exporter

Rather than passing an exporter to every `with_trace()` call, set a
session-wide default. This is especially useful in production scripts where
every trace should go to the same destination.

```{r}
trace_file <- tempfile(fileext = ".jsonl")
set_default_exporter(jsonl_exporter(trace_file))

# All subsequent traces are automatically exported
with_trace("auto-exported-1", {
  with_span("work", type = "tool", { 1 + 1 })
})

with_trace("auto-exported-2", {
  with_span("more-work", type = "tool", { 2 + 2 })
})

lines <- readLines(trace_file)
length(lines)  # 2 traces exported

unlink(trace_file)
```

## Custom Exporters

The built-in exporters cover the most common cases, but you can build your own
for any destination. Pass a function to `new_exporter()` -- the function
receives the serialized trace list (from `$to_list()`) as its single argument.

This makes it straightforward to integrate with databases, message queues,
HTTP APIs, or any custom logging system:

```{r}
# A simple exporter that counts spans per trace
span_counter <- new_exporter(function(trace_list) {
  cat(sprintf(
    "Trace '%s' completed with %d spans\n",
    trace_list$name,
    length(trace_list$spans)
  ))
})

with_trace("custom-export", exporter = span_counter, {
  with_span("a", type = "tool", { 1 })
  with_span("b", type = "tool", { 2 })
  with_span("c", type = "llm", {
    record_tokens(100, 50, model = "claude-haiku-4-5")
    3
  })
})
```

## Serialization

Every trace and span can be serialized to a plain R list with `$to_list()`.
This is what exporters receive internally, and it is the right format for
programmatic analysis -- you get standard R data structures instead of R6
reference objects.

```{r}
tr <- Trace$new("serialize-demo")
tr$start()

s <- Span$new("work", type = "custom")
s$start()
s$add_metric("quality", 0.95)
s$end()
tr$add_span(s)

tr$end()

trace_list <- tr$to_list()
names(trace_list)
trace_list$name
trace_list$spans[[1]]$metrics
```

## trace_schema() -- JSONL Format Reference

When building parsers, dashboards, or validation pipelines for exported trace
data, you need to know exactly what fields to expect. `trace_schema()` returns
a list describing every field in the JSONL export format, including types and
descriptions for both trace-level and span-level fields:

```{r}
schema <- trace_schema()
names(schema)
#> [1] "trace_id" "name" "status" "start_time" "end_time" "duration" "spans"

# Inspect span-level fields
names(schema$spans$fields)
#> [1] "span_id" "name" "type" "status" "start_time" "end_time"
#> [7] "duration_secs" "parent_id" "model" "input_tokens" "output_tokens"
```

## Next Steps

- `vignette("cloud-native")` -- OTLP export to Jaeger/Tempo, Prometheus
  metrics, and W3C Trace Context propagation for distributed tracing.
- `vignette("orchestr-integration")` -- Automatic tracing of orchestr graph
  executions and agent invocations.
