---
title: "Exporters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exporters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## JSONL exporter

Write each trace as one JSON line to an append-only file.

```{r}
library(securetrace)
trace_file <- tempfile(fileext = ".jsonl")
exp <- exporter_jsonl(trace_file)
with_trace("exported-run", exporter = exp, {
  with_span("llm-call", type = "llm", {
    record_tokens(1500, 300, model = "claude-sonnet-4-5")
    "response"
  })
  with_span("tool-call", type = "tool", { 42 })
})
```

Read back with `jsonlite`:

```{r}
lines <- readLines(trace_file)
length(lines)
trace_data <- jsonlite::fromJSON(lines[[1]])
trace_data$name
length(trace_data$spans$span_id)
unlink(trace_file)
```

## Console exporter

Print trace summaries during interactive development.

```{r}
debug_exp <- exporter_console(verbose = TRUE)
with_trace("debug-run", exporter = debug_exp, {
  with_span("planning", type = "llm", {
    record_tokens(2000, 500, model = "claude-sonnet-4-5")
    "plan"
  })
  with_span("execution", type = "tool", { 100 })
})
```

Set `verbose = FALSE` for headers only.

## Multiple exporters

Fan out to N destinations with `exporter_multi()`.

```{r}
trace_file <- tempfile(fileext = ".jsonl")
combined <- exporter_multi(
  exporter_jsonl(trace_file),
  exporter_console(verbose = TRUE)
)
with_trace("multi-export-run", exporter = combined, {
  with_span("llm-call", type = "llm", {
    record_tokens(3000, 600, model = "claude-opus-4-6")
    "result"
  })
})
readLines(trace_file) |> length()
unlink(trace_file)
```

## Default exporter

Set a session-wide exporter instead of passing one per `with_trace()` call.

```{r}
trace_file <- tempfile(fileext = ".jsonl")
set_default_exporter(exporter_jsonl(trace_file))
with_trace("auto-1", { with_span("work", type = "tool", { 1 + 1 }) })
with_trace("auto-2", { with_span("more-work", type = "tool", { 2 + 2 }) })
readLines(trace_file) |> length()
unlink(trace_file)
```

## Custom exporters

Pass any function to `exporter()`. It receives the serialized trace list.

```{r}
span_counter <- exporter(function(trace_list) {
  cat(sprintf("Trace '%s': %d spans\n",
              trace_list$name, length(trace_list$spans)))
})
with_trace("custom-export", exporter = span_counter, {
  with_span("a", type = "tool", { 1 })
  with_span("b", type = "tool", { 2 })
  with_span("c", type = "llm", {
    record_tokens(100, 50, model = "claude-haiku-4-5")
    3
  })
})
```

## Serialization

Convert any trace to a plain list with `$to_list()` -- this is what exporters receive.

```{r}
tr <- Trace$new("serialize-demo")
tr$start()
s <- Span$new("work", type = "custom")
s$start()
s$add_metric("quality", 0.95)
s$end()
tr$add_span(s)
tr$end()
trace_list <- tr$to_list()
names(trace_list)
trace_list$spans[[1]]$metrics
```

Discover all JSONL fields with `trace_schema()`:

```{r}
schema <- trace_schema()
names(schema)
names(schema$spans$fields)
```
