---
title: "Traces, Spans, and Events"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Traces, Spans, and Events}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The R6 API gives you direct access to events, custom metrics, nested spans, error handling, and integration helpers. For `with_trace()` basics and token tracking, see `vignette("securetrace")`.

```{r}
library(securetrace)
```

## Span types

| Type | Use for | Example |
|------|---------|---------|
| `"llm"` | Language model calls | `Span$new("plan", type = "llm")` |
| `"tool"` | Tool / function execution | `Span$new("calc", type = "tool")` |
| `"guardrail"` | Input/output validation | `Span$new("pii-check", type = "guardrail")` |
| `"custom"` | Anything else | `Span$new("transform", type = "custom")` |

## Nested spans

Inner spans record the outer span's ID as `parent_id`:

```
Trace: "pipeline"
|-- Span: "planning" (llm)
|   '-- Span: "calculator" (tool)   <- child
'-- Span: "summarize" (llm)         <- sibling
```

```{r}
result <- with_trace("pipeline", {
  with_span("planning", type = "llm", {
    record_tokens(2000, 500, model = "claude-sonnet-4-5")
    with_span("calculator", type = "tool", { 2 + 2 })
  })
  with_span("summarize", type = "llm", {
    record_tokens(1000, 200, model = "claude-haiku-4-5")
    "done"
  })
})
```

Parent-child relationships persist in exported JSON via `parent_id` fields.

## Manual construction

Use the R6 classes when spans are created in one function and ended in another, or when building traces from recorded data.

```{r}
tr <- Trace$new("manual-trace", metadata = list(user = "analyst"))
tr$start()
s1 <- Span$new("llm-call", type = "llm")
s1$start()
s1$set_model("claude-opus-4-6")
s1$set_tokens(input = 5000L, output = 1000L)
s1$end()
tr$add_span(s1)
s2 <- Span$new("tool-use", type = "tool", parent_id = s1$span_id)
s2$start()
s2$add_metric("rows_processed", 150, unit = "rows")
s2$end()
tr$add_span(s2)
tr$end()
```

```{r}
tr$status
tr$duration()
length(tr$spans)
```

## Events

Create with `trace_event()`, attach with `$add_event()`, access with `@`:

```{r}
tr <- Trace$new("event-demo")
tr$start()
s <- Span$new("llm-call", type = "llm")
s$start()
s$add_event(trace_event("model_selected", data = list(model = "claude-sonnet-4-5")))
s$add_event(trace_event("prompt_sent", data = list(length = 1500L)))
s$set_tokens(input = 1500L, output = 300L)
s$end()
tr$add_span(s)
tr$end()
```

```{r}
length(s$events)
s$events[[1]]@name
s$events[[1]]@data
```

## Custom metrics

Use `record_metric()` / `record_latency()` inside `with_span()`, or `$add_metric()` on the R6 object:

```{r}
result <- with_trace("metrics-demo", {
  with_span("data-processing", type = "tool", {
    record_metric("rows_processed", 1500, unit = "rows")
    record_latency(0.42)
    "processed"
  })
})
```

```{r}
s <- Span$new("manual-metrics", type = "tool")
s$start()
s$add_metric("cache_hits", 42, unit = "count")
s$end()
s$metrics
```

## Error handling

Errors inside `with_span()` set status to `"error"` and record the message. The trace still exports, then re-raises.

```{r, error = TRUE}
trace_file <- tempfile(fileext = ".jsonl")
exp <- jsonl_exporter(trace_file)
with_trace("error-run", exporter = exp, {
  with_span("failing-step", type = "tool", {
    stop("something went wrong")
  })
})
```

```{r}
lines <- readLines(trace_file)
trace_data <- jsonlite::fromJSON(lines[[1]])
trace_data$status
trace_data$spans$status
trace_data$spans$error
unlink(trace_file)
```

## Integration helpers

`trace_tool_call()` and `trace_guardrail()` create typed spans automatically. Both require an active `with_trace()`.

```{r}
with_trace("tool-integration", {
  trace_tool_call("calculator", function(x) x * 2, 21)
})
```

```{r}
with_trace("guard-integration", {
  trace_guardrail("length-check", function(x) nchar(x) < 1000, "short text")
})
```

### Secure execution

`trace_execution()` wraps `securer::SecureSession$execute()` with a span, recording code and stdout as events:

```{r, eval = FALSE}
session <- securer::SecureSession$new()
with_trace("sandboxed-run", {
  trace_execution(session, "cat('hello'); 1 + 1")
})
session$close()
```

### secureguard integration

Pass a Guard object to `trace_guardrail()` for structured result metadata:

```{r, eval = FALSE}
guard <- secureguard::guard_code_analysis()
with_trace("guarded-input", {
  trace_guardrail("code-safety", guard, "system('rm -rf /')")
})
```

## Full example

Multi-step workflow: nested spans, tokens, events, metrics, and JSONL export.

```{r}
trace_file <- tempfile(fileext = ".jsonl")
exp <- multi_exporter(jsonl_exporter(trace_file), console_exporter(verbose = TRUE))
result <- with_trace("full-workflow", exporter = exp, {
  with_span("planning", type = "llm", {
    record_tokens(3000, 800, model = "claude-sonnet-4-5")
    current_span()$add_event(trace_event("model_selected",
      data = list(model = "claude-sonnet-4-5")))
    list(steps = c("fetch", "compute", "summarize"))
  })
  data <- with_span("fetch-data", type = "tool", {
    record_metric("rows_fetched", 250, unit = "rows")
    data.frame(x = 1:250, y = rnorm(250))
  })
  with_span("compute", type = "custom", {
    with_span("validate", type = "guardrail", { nrow(data) > 0 })
    with_span("transform", type = "tool", { mean(data$y) })
  })
  with_span("summarize", type = "llm", {
    record_tokens(1500, 400, model = "claude-haiku-4-5")
    sprintf("Processed %d rows, mean = %.2f", nrow(data), mean(data$y))
  })
})
lines <- readLines(trace_file)
trace_data <- jsonlite::fromJSON(lines[[1]])
sprintf("Trace '%s': %d spans, status = %s",
        trace_data$name, length(trace_data$spans$span_id), trace_data$status)
unlink(trace_file)
```

## Next steps

- `vignette("exporters")`: JSONL, console, custom exporters, and trace schema.
- `vignette("cloud-native")`: OTLP, Prometheus, W3C Trace Context.
- `vignette("orchestr-integration")`: automatic tracing of orchestr graphs.
