---
title: "Observability Integration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Observability Integration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

securetrace provides structured observability for R-based LLM agent
workflows. This vignette covers how to instrument your agents with traces
and spans, record token and cost data, and export traces for analysis.

## Core Concepts

securetrace follows an OpenTelemetry-inspired model with three primitives:

- **Trace** -- A root container representing a complete agent run. Each trace
  has a unique ID, a name, metadata, and a collection of spans.
- **Span** -- A single operation within a trace. Spans have types (`"llm"`,
  `"tool"`, `"guardrail"`, `"custom"`), can record token usage, and can nest
  inside each other via parent IDs.
- **Event** -- A discrete point-in-time occurrence within a span, such as a
  model selection or a prompt being sent.

## Creating Traces and Spans

The simplest way to instrument code is with the `with_trace()` and
`with_span()` context managers:

```{r}
library(securetrace)

result <- with_trace("my-agent-run", {
  with_span("planning", type = "llm", {
    record_tokens(1500, 300, model = "claude-sonnet-4-5")
    "The plan is ready"
  })
})

result
```

`with_trace()` creates a `Trace` object, starts it, evaluates the expression,
ends the trace, and optionally exports it. Inside the trace, `with_span()`
creates child spans the same way. The current trace and span are available via
`current_trace()` and `current_span()` at any point during evaluation.

## Span Nesting

Spans can nest to represent hierarchical operations. When you call
`with_span()` inside another `with_span()`, the inner span automatically
records the outer span's ID as its `parent_id`:

```{r}
result <- with_trace("nested-workflow", {

  # Outer span: planning step

  with_span("planning", type = "llm", {
    record_tokens(2000, 500, model = "claude-sonnet-4-5")

    # Inner span: a tool call within the planning step
    with_span("calculator", type = "tool", {
      2 + 2
    })
  })

  # Sibling span: summarization step
  with_span("summarize", type = "llm", {
    record_tokens(1000, 200, model = "claude-haiku-4-5")
    "Summary complete"
  })

  # Capture the trace while still inside with_trace()
  tr <- current_trace()
})
```

Each span knows its parent. When the trace is serialized with `$to_list()`,
the parent-child relationships are preserved through `parent_id` fields,
letting you reconstruct the full call tree.

## Manual Trace and Span Construction

For full control, you can use the R6 classes directly instead of the
context-manager wrappers:
```{r}
# Create a trace manually
tr <- Trace$new("manual-trace", metadata = list(user = "analyst"))
tr$start()

# Create and configure a span
s1 <- Span$new("llm-call", type = "llm")
s1$start()
s1$set_model("claude-opus-4-6")
s1$set_tokens(input = 5000L, output = 1000L)
s1$end()
tr$add_span(s1)

# Create a second span with a parent
s2 <- Span$new("tool-use", type = "tool", parent_id = s1$span_id)
s2$start()
s2$add_metric("rows_processed", 150, unit = "rows")
s2$end()
tr$add_span(s2)

tr$end()

tr$status
tr$duration()
length(tr$spans)
```

## Recording Events

Events capture discrete occurrences within a span. They are S7 objects
created with `trace_event()` and attached to spans with `$add_event()`:

```{r}
tr <- Trace$new("event-demo")
tr$start()

s <- Span$new("llm-call", type = "llm")
s$start()

# Record events during the span
evt1 <- trace_event("model_selected", data = list(model = "claude-sonnet-4-5"))
s$add_event(evt1)

evt2 <- trace_event("prompt_sent", data = list(length = 1500L))
s$add_event(evt2)

s$set_tokens(input = 1500L, output = 300L)
s$end()
tr$add_span(s)
tr$end()

# Events are accessible on the span
length(s$events)
s$events[[1]]@name
s$events[[1]]@data
```

Note the `@` accessor for S7 properties on events (and exporters), versus `$`
for R6 fields on Trace and Span objects.

## Token and Cost Accounting

### Recording Tokens

Inside a `with_span()` block, use `record_tokens()` to log token usage on the
current span:

```{r}
result <- with_trace("token-tracking", {
  with_span("step-1", type = "llm", {
    record_tokens(2000, 500, model = "claude-sonnet-4-5")
    "Step 1 done"
  })

  with_span("step-2", type = "llm", {
    record_tokens(1000, 200, model = "claude-haiku-4-5")
    "Step 2 done"
  })
})
```

When using the R6 API directly, call `$set_tokens()` and `$set_model()` on the
span object:

```{r}
s <- Span$new("direct-span", type = "llm")
s$start()
s$set_tokens(input = 3000L, output = 800L)
s$set_model("gpt-4o")
s$end()
```

### Built-in Model Pricing

securetrace ships with per-1M-token pricing for common models:

```{r}
# View all known model prices
costs <- model_costs()
names(costs)
costs[["claude-sonnet-4-5"]]
```

### Calculating Costs

Use `calculate_cost()` for a single model call, or `trace_total_cost()` for
an entire trace:

```{r}
# Cost of a single call
calculate_cost("claude-opus-4-6", input_tokens = 5000, output_tokens = 1000)

# Build a trace with multiple LLM spans
tr <- Trace$new("cost-demo")
tr$start()

s1 <- Span$new("expensive-call", type = "llm")
s1$start()
s1$set_model("claude-opus-4-6")
s1$set_tokens(input = 10000L, output = 2000L)
s1$end()
tr$add_span(s1)

s2 <- Span$new("cheap-call", type = "llm")
s2$start()
s2$set_model("claude-haiku-4-5")
s2$set_tokens(input = 5000L, output = 1000L)
s2$end()
tr$add_span(s2)

tr$end()

# Total cost across all spans
trace_total_cost(tr)
```

### Custom Model Pricing

Register pricing for models not in the built-in table:

```{r}
add_model_cost("my-fine-tuned-model", input_price = 5, output_price = 20)
calculate_cost("my-fine-tuned-model", input_tokens = 1000, output_tokens = 500)
```

### Custom Metrics

Record arbitrary metrics on spans with `record_metric()` (inside a
`with_span()`) or `$add_metric()` (on the R6 object):

```{r}
result <- with_trace("metrics-demo", {
  with_span("data-processing", type = "tool", {
    record_metric("rows_processed", 1500, unit = "rows")
    record_latency(0.42)
    "processed"
  })
})
```

## Trace Summary

The `$summary()` method prints a formatted overview of a completed trace,
including span count, total tokens, and estimated cost:

```{r}
tr <- Trace$new("summary-demo")
tr$start()

s <- Span$new("llm-call", type = "llm")
s$start()
s$set_model("claude-opus-4-6")
s$set_tokens(input = 5000L, output = 1000L)
s$end()
tr$add_span(s)

tr$end()
tr$summary()
```

## JSONL Export

The `jsonl_exporter()` writes each completed trace as a single JSON line to
a file. This is ideal for post-hoc analysis, dashboards, or feeding into
observability platforms:

```{r}
# Create a temporary file for this example
trace_file <- tempfile(fileext = ".jsonl")

exp <- jsonl_exporter(trace_file)

# Traces are exported automatically when with_trace() completes
with_trace("exported-run", exporter = exp, {
  with_span("llm-call", type = "llm", {
    record_tokens(1500, 300, model = "claude-sonnet-4-5")
    "response"
  })

  with_span("tool-call", type = "tool", {
    42
  })
})

# Read back the exported data
lines <- readLines(trace_file)
length(lines)  # 1 line per trace

# Parse and inspect
trace_data <- jsonlite::fromJSON(lines[[1]])
trace_data$name
trace_data$status
length(trace_data$spans$span_id)

# Clean up
unlink(trace_file)
```

Each JSONL line contains the full trace structure: trace ID, name, status,
timestamps, duration, and all spans with their tokens, events, and metrics.

## Console Exporter

The `console_exporter()` prints trace summaries directly to the console,
useful for debugging and interactive development:

```{r}
debug_exp <- console_exporter(verbose = TRUE)

with_trace("debug-run", exporter = debug_exp, {
  with_span("planning", type = "llm", {
    record_tokens(2000, 500, model = "claude-sonnet-4-5")
    "plan"
  })

  with_span("execution", type = "tool", {
    100
  })
})
```

Set `verbose = FALSE` to print only the trace header without individual span
details.

## Multi-Exporter Setup

Use `multi_exporter()` to send traces to multiple destinations at once.
A common pattern is logging to both a file and the console:

```{r}
trace_file <- tempfile(fileext = ".jsonl")

# Combine a file exporter and a console exporter
file_exp <- jsonl_exporter(trace_file)
console_exp <- console_exporter(verbose = TRUE)
combined <- multi_exporter(file_exp, console_exp)

with_trace("multi-export-run", exporter = combined, {
  with_span("llm-call", type = "llm", {
    record_tokens(3000, 600, model = "claude-opus-4-6")
    "result"
  })
})

# The trace was written to the file AND printed to console
lines <- readLines(trace_file)
length(lines)

unlink(trace_file)
```

## Default Exporter

Rather than passing an exporter to every `with_trace()` call, set a
session-wide default:

```{r}
trace_file <- tempfile(fileext = ".jsonl")
set_default_exporter(jsonl_exporter(trace_file))

# All subsequent traces are automatically exported
with_trace("auto-exported-1", {
  with_span("work", type = "tool", { 1 + 1 })
})

with_trace("auto-exported-2", {
  with_span("more-work", type = "tool", { 2 + 2 })
})

lines <- readLines(trace_file)
length(lines)  # 2 traces exported

unlink(trace_file)
```

## Custom Exporters

Build your own exporter by passing a function to `new_exporter()`. The
function receives the serialized trace list (from `$to_list()`) as its
single argument:

```{r}
# A simple exporter that counts spans per trace
span_counter <- new_exporter(function(trace_list) {
  cat(sprintf(
    "Trace '%s' completed with %d spans\n",
    trace_list$name,
    length(trace_list$spans)
  ))
})

with_trace("custom-export", exporter = span_counter, {
  with_span("a", type = "tool", { 1 })
  with_span("b", type = "tool", { 2 })
  with_span("c", type = "llm", {
    record_tokens(100, 50, model = "claude-haiku-4-5")
    3
  })
})
```

## Error Handling

Traces and spans capture errors gracefully. When an error occurs inside
`with_span()`, the span's status is set to `"error"` and the error message
is recorded. The error then propagates to the enclosing `with_trace()`,
which marks itself as `"error"` and exports before re-raising:

```{r, error = TRUE}
trace_file <- tempfile(fileext = ".jsonl")
exp <- jsonl_exporter(trace_file)

with_trace("error-run", exporter = exp, {
  with_span("failing-step", type = "tool", {
    stop("something went wrong")
  })
})
```

```{r}
# The trace was still exported despite the error
lines <- readLines(trace_file)
trace_data <- jsonlite::fromJSON(lines[[1]])
trace_data$status
trace_data$spans$status
trace_data$spans$error

unlink(trace_file)
```

## Serialization

Every trace and span can be serialized to a plain R list with `$to_list()`.
This is what exporters receive, and it is useful for programmatic analysis:

```{r}
tr <- Trace$new("serialize-demo")
tr$start()

s <- Span$new("work", type = "custom")
s$start()
s$add_metric("quality", 0.95)
s$end()
tr$add_span(s)

tr$end()

trace_list <- tr$to_list()
names(trace_list)
trace_list$name
trace_list$spans[[1]]$metrics
```

## Integration Helpers

securetrace provides convenience wrappers for common instrumentation
patterns. These require `with_trace()` to be active:

```{r}
# Wrap tool calls
result <- with_trace("tool-integration", {
  trace_tool_call("calculator", function(x) x * 2, 21)
})
result

# Wrap guardrail checks
result <- with_trace("guard-integration", {
  trace_guardrail("length-check", function(x) nchar(x) < 1000, "short text")
})
result
```

For `trace_llm_call()` (wrapping ellmer chat objects) and `trace_execution()`
(wrapping securer sessions), see the package documentation -- these require
their respective optional dependencies.

## Putting It All Together

Here is a complete example of an instrumented multi-step agent workflow with
file export:

```{r}
trace_file <- tempfile(fileext = ".jsonl")

combined_exp <- multi_exporter(
  jsonl_exporter(trace_file),
  console_exporter(verbose = TRUE)
)

result <- with_trace("full-agent-workflow", exporter = combined_exp, {

  # Step 1: Planning (LLM call)
  plan <- with_span("planning", type = "llm", {
    record_tokens(3000, 800, model = "claude-sonnet-4-5")
    evt <- trace_event("model_selected", data = list(model = "claude-sonnet-4-5"))
    current_span()$add_event(evt)
    list(steps = c("fetch", "compute", "summarize"))
  })

  # Step 2: Tool execution
  data <- with_span("fetch-data", type = "tool", {
    record_metric("rows_fetched", 250, unit = "rows")
    record_latency(0.15)
    data.frame(x = 1:250, y = rnorm(250))
  })

  # Step 3: Nested computation
  with_span("compute", type = "custom", {

    with_span("validate", type = "guardrail", {
      nrow(data) > 0
    })

    with_span("transform", type = "tool", {
      mean(data$y)
    })
  })

  # Step 4: Summarization (LLM call)
  with_span("summarize", type = "llm", {
    record_tokens(1500, 400, model = "claude-haiku-4-5")
    sprintf("Processed %d rows, mean = %.2f", nrow(data), mean(data$y))
  })
})

result

# Verify export
lines <- readLines(trace_file)
trace_data <- jsonlite::fromJSON(lines[[1]])
sprintf("Trace '%s': %d spans, status = %s",
        trace_data$name,
        length(trace_data$spans$span_id),
        trace_data$status)

unlink(trace_file)
```
