---
title: "Tracing orchestr Workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tracing orchestr Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Setup

securetrace integrates with [orchestr](https://github.com/ian-flores/orchestr)
to automatically trace graph executions and agent invocations. Install both
packages to use these features:

```{r}
library(securetrace)
library(orchestr)
```

## Tracing a Graph

`trace_graph()` wraps `graph$invoke()` and creates a child span for each node
that executes. Build a graph with `graph_builder()`, then pass the compiled
graph and initial state:

```{r}
gb <- graph_builder()
gb$add_node("fetch", function(state, config) list(data = mtcars[1:5, ]))
gb$add_node("summarize", function(state, config) list(summary = summary(state$data)))
gb$add_edge("fetch", "summarize")
gb$add_edge("summarize", END)
gb$set_entry_point("fetch")
graph <- gb$compile()

result <- trace_graph(graph, list(data = NULL, summary = NULL))
```

Each node execution is captured as a span named `"node:{name}"`, with metadata
recording the node name and step number.

## Tracing an Agent

`trace_agent()` wraps a single `agent$invoke()` call. It auto-extracts the
model name and token usage from the underlying ellmer Chat object:

```{r}
chat <- ellmer::chat_openai(model = "gpt-4o")
assistant <- agent("assistant", chat)

response <- trace_agent(assistant, "Summarize the iris dataset.")
```

Token deltas are computed by comparing the chat's token state before and after
the invocation, so multi-turn agents report only the tokens used per call.

## Combining with Exporters

Pass an exporter to `trace_graph()` to persist the full trace as JSONL:

```{r}
exp <- jsonl_exporter("graph-traces.jsonl")
result <- trace_graph(graph, list(data = NULL, summary = NULL), exporter = exp)

# Or set a default exporter for all traces
set_default_exporter(exp)
result <- trace_graph(graph, list(data = NULL, summary = NULL))
```

## Cost Tracking

When graph nodes make LLM calls that record model and token information,
`trace_total_cost()` aggregates costs across all spans:

```{r}
chat <- ellmer::chat_openai(model = "gpt-4o")
graph <- react_graph(agent("analyst", chat))

result <- trace_graph(graph, list(messages = list("Analyze sales trends.")))
tr <- current_trace()
trace_total_cost(tr)
#> [1] 0.00045
```

## Manual Instrumentation

For finer control, use `with_trace()` and `with_span()` directly inside
graph node handlers instead of `trace_graph()`:

```{r}
analyze_node <- function(state, config) {
  with_span("llm-call", type = "llm", {
    record_tokens(500, 120, model = "claude-sonnet-4-5")
    list(result = "analysis complete")
  })
}

gb <- graph_builder()
gb$add_node("analyze", analyze_node)
gb$add_edge("analyze", END)
gb$set_entry_point("analyze")
graph <- gb$compile()

with_trace("custom-graph-run", {
  graph$invoke(list(result = NULL))
})
```

This lets you control span names, types, and metadata at each step, and is
useful when you need to trace sub-operations within a single node.
