---
title: "Tracing orchestr Workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tracing orchestr Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
has_orchestr <- requireNamespace("orchestr", quietly = TRUE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = has_orchestr)
```

```{r}
library(securetrace)
library(orchestr)
```

## trace_graph()

Auto-instrument every node in a compiled graph: one span per node, zero handler changes.

```{r}
gb <- graph_builder()
gb$add_node("fetch", function(state, config) list(data = mtcars[1:5, ]))
gb$add_node("summarize", function(state, config) list(summary = summary(state$data)))
gb$add_edge("fetch", "summarize")
gb$add_edge("summarize", END)
gb$set_entry_point("fetch")
graph <- gb$compile()

result <- trace_graph(graph, list(data = NULL, summary = NULL))
```

## trace_agent()

Wrap a single agent invocation; model name and token deltas are extracted from the ellmer chat object automatically.

```{r, eval = FALSE}
chat <- ellmer::chat_openai(model = "gpt-4o")
assistant <- agent("assistant", chat)
response <- trace_agent(assistant, "Summarize the iris dataset.")
```

## With exporters

Pass an exporter to persist the trace as JSONL.

```{r}
exp <- exporter_jsonl(tempfile(fileext = ".jsonl"))
result <- trace_graph(graph, list(data = NULL, summary = NULL), exporter = exp)

# Or set a default for all traces
set_default_exporter(exporter_jsonl(tempfile(fileext = ".jsonl")))
result <- trace_graph(graph, list(data = NULL, summary = NULL))
```

## Cost tracking

Aggregate costs across all spans. Useful for ReAct loops where call count is unpredictable.

```{r, eval = FALSE}
chat <- ellmer::chat_openai(model = "gpt-4o")
graph <- react_graph(agent("analyst", chat))
result <- trace_graph(graph, list(messages = list("Analyze sales trends.")))
trace_total_cost(current_trace())
```

## Manual spans in nodes

Add `with_span()` inside a handler for sub-node detail. Combine with `trace_graph()` freely.

```{r}
analyze_node <- function(state, config) {
  with_span("llm-call", type = "llm", {
    record_tokens(500, 120, model = "claude-sonnet-4-5")
    list(result = "analysis complete")
  })
}

gb <- graph_builder()
gb$add_node("analyze", analyze_node)
gb$add_edge("analyze", END)
gb$set_entry_point("analyze")

with_trace("custom-graph-run", {
  gb$compile()$invoke(list(result = NULL))
})
```
