---
title: "Getting Started with securetrace"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with securetrace}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

securetrace provides observability and tracing for R-based LLM agent workflows.
It tracks structured traces with spans, records token and cost accounting,
monitors latency, and exports data as JSONL for analysis.

## Basic Usage

Create a trace with spans to track an agent workflow:

```{r}
library(securetrace)

# Wrap your workflow in a trace
result <- with_trace("my-agent-run", {

  # Each operation gets its own span
  with_span("llm-call", type = "llm", {
    record_tokens(1500, 300, model = "claude-sonnet-4-5")
    "The answer is 42"
  })
})
```

## Token and Cost Tracking

securetrace includes built-in pricing for popular LLM models:

```{r}
# View known model pricing (per 1M tokens)
model_costs()

# Calculate cost for a specific call
calculate_cost("claude-opus-4-6", input_tokens = 5000, output_tokens = 1000)

# Register custom model pricing
add_model_cost("my-fine-tuned-model", input_price = 5, output_price = 20)
```

### Multi-Provider Support

securetrace ships with pricing for models across Anthropic, OpenAI, Google
Gemini, Mistral, and DeepSeek:

```{r}
# View all known model pricing (per 1M tokens)
costs <- model_costs()
head(costs)
#>                  model input_price output_price
#> 1     claude-opus-4-6       15.00        75.00
#> 2   claude-sonnet-4-5        3.00        15.00
#> 3              gpt-4o        2.50        10.00
#> 4      gemini-1.5-pro        1.25         5.00
#> 5      gemini-2.0-flash      0.10         0.40
#> 6  deepseek-reasoner        0.55         2.19
```

Cloud provider model IDs (AWS Bedrock, Google Vertex) are resolved
automatically via built-in aliases:

```{r}
# Bedrock model ID works transparently
calculate_cost(
  "anthropic.claude-3-5-sonnet-20241022-v2:0",
  input_tokens = 10000,
  output_tokens = 2000
)

# Vertex model ID also resolves
calculate_cost(
  "publishers/anthropic/models/claude-3-5-sonnet-v2@20241022",
  input_tokens = 10000,
  output_tokens = 2000
)
```

For custom or self-hosted deployments, register your own aliases:

```{r}
# Map an internal deployment name to a known model
add_model_alias("my-company-claude", "claude-sonnet-4-5")
calculate_cost("my-company-claude", input_tokens = 5000, output_tokens = 1000)
```

When using `trace_llm_call()` with an ellmer Chat object, model name and token
counts are extracted automatically:

```{r}
result <- with_trace("auto-instrumented", {
  trace_llm_call("analysis", chat, "Summarize the data.")
})
# Model and token counts are set on the span without manual recording
```

## Exporting Traces

Export completed traces to JSONL files or the console:
```{r}
# JSONL file export
exp <- jsonl_exporter("traces.jsonl")

with_trace("exported-run", exporter = exp, {
  with_span("work", type = "tool", {
    42
  })
})

# Console export for debugging
debug_exp <- console_exporter(verbose = TRUE)

# Combine multiple exporters
both <- multi_exporter(exp, debug_exp)

# Or set a default exporter for all traces
set_default_exporter(exp)
```

## Integration Helpers

securetrace provides helpers for common operations:

```{r}
# Wrap tool calls
result <- with_trace("tools", {
  trace_tool_call("calculator", function(x) x * 2, 21)
})

# Wrap guardrail checks
with_trace("guarded", {
  trace_guardrail("length-check", function(x) nchar(x) < 1000, user_input)
})
```

### Streaming LLM Calls

`trace_llm_call()` supports streaming responses via the `stream` parameter.
When `stream = TRUE`, the chat object's `$stream()` method is called instead
of `$chat()`, and a `streaming` event is recorded on the span:

```{r}
# Stream responses with automatic tracing
with_trace("streaming-example", {
  response <- trace_llm_call(chat, "Explain R6 classes", stream = TRUE)
})
# The span records: model, tokens, latency, and a "streaming" event
```

### Tool Call Tracking

When the LLM invokes tools during a chat, `trace_llm_call()` automatically
inspects the response turn and records each tool invocation as a `tool_call`
event on the span. No extra instrumentation is needed:

```{r}
# Register tools on the chat object, then trace the call
chat <- ellmer::chat_openai(model = "gpt-4o")
chat$register_tool(ellmer::tool(
  function(x, y) x + y,
  "Adds two numbers", x = "First number", y = "Second number"
))

with_trace("agent-with-tools", {
  response <- trace_llm_call(chat, "What is 3 + 4?")
})
# Each tool invocation appears as a "tool_call" event with name and arguments
```

## Nested Spans

Spans can be nested to represent hierarchical operations:

```{r}
with_trace("complex-workflow", {
  with_span("planning", type = "llm", {
    record_tokens(2000, 500, model = "claude-sonnet-4-5")

    with_span("tool-use", type = "tool", {
      # Tool execution within the planning step
      42
    })
  })

  with_span("summarize", type = "llm", {
    record_tokens(1000, 200, model = "claude-haiku-4-5")
    "Summary complete"
  })
})
```

## Trace Summary

Get a formatted summary of any trace:

```{r}
tr <- Trace$new("manual-trace")
tr$start()
s <- Span$new("llm", type = "llm")
s$start()
s$set_tokens(input = 5000, output = 1000)
s$set_model("claude-opus-4-6")
s$end()
tr$add_span(s)
tr$end()

tr$summary()
# Trace: manual-trace (completed)
#   Duration: 0.01s
#   Spans: 1
#   Tokens: 5000 input, 1000 output
#   Cost: $0.150000
```
