---
title: "Getting Started with securetrace"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with securetrace}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

securetrace gives you structured tracing, token accounting, and cost tracking
for LLM agent workflows in R.

## Quick start

Wrap your workflow in `with_trace()`, break it into spans, and record tokens:

```{r}
library(securetrace)

result <- with_trace("my-agent-run", {
  with_span("planning", type = "llm", {
    record_tokens(1500, 300, model = "claude-sonnet-4-5")
    "The answer is 42"
  })
})
```

- `with_trace()` creates a `Trace`, starts the clock, evaluates your code, and
  ends the trace.
- `with_span()` wraps a single operation (LLM call, tool use, etc.).
- `record_tokens()` logs input/output tokens and model on the current span.

Use `current_trace()` and `current_span()` to access active objects anywhere
inside the block.

## Token and cost tracking

Built-in pricing covers Anthropic, OpenAI, Gemini, Mistral, and DeepSeek.

```{r}
# All known pricing (per 1M tokens)
costs <- model_costs()
head(names(costs))
```

```{r}
# Cost for a single call
calculate_cost("claude-sonnet-4-5", input_tokens = 5000, output_tokens = 1000)
```

Register your own models:

```{r}
add_model_cost("my-fine-tuned", input_price = 5, output_price = 20)
calculate_cost("my-fine-tuned", input_tokens = 10000, output_tokens = 2000)
```

Cloud provider model IDs (Bedrock, Vertex) resolve automatically via
`resolve_model()`:

```{r}
calculate_cost(
  "anthropic.claude-3-5-sonnet-20241022-v2:0",
  input_tokens = 10000, output_tokens = 2000
)
```

Map internal deployment names with `add_model_alias()`:

```{r}
add_model_alias("my-company-claude", "claude-sonnet-4-5")
calculate_cost("my-company-claude", input_tokens = 5000, output_tokens = 1000)
```

## Exporting traces

Write traces to JSONL for downstream analysis:

```{r}
exp <- jsonl_exporter(tempfile("traces", fileext = ".jsonl"))

with_trace("exported-run", exporter = exp, {
  with_span("work", type = "tool", { 42 })
})
```

Print to console while debugging:

```{r}
debug_exp <- console_exporter(verbose = TRUE)

with_trace("debug-run", exporter = debug_exp, {
  with_span("step", type = "custom", { 1 + 1 })
})
```

Set a default exporter so every `with_trace()` auto-exports:

```{r}
set_default_exporter(exp)
```

See `vignette("exporters")` for custom exporters, `multi_exporter()`, and the
JSONL schema reference.

## Trace summary

Call `$summary()` on a completed trace to see duration, span count, tokens, and
cost at a glance:

```{r}
tr <- Trace$new("summarized-run")
tr$start()
s <- Span$new("llm", type = "llm")
s$start()
s$set_tokens(input = 5000, output = 1000)
s$set_model("claude-opus-4-6")
s$end()
tr$add_span(s)
tr$end()

tr$summary()
```

## Next steps

- `vignette("observability")` -- spans, events, metrics, error handling, nested workflows.
- `vignette("exporters")` -- JSONL schema, console exporter, custom exporters.
- `vignette("cloud-native")` -- OTLP, Prometheus, W3C Trace Context.
