---
title: "Getting Started with securetrace"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with securetrace}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

securetrace provides observability and tracing for R-based LLM agent workflows.
It tracks structured traces with spans, records token and cost accounting,
monitors latency, and exports data as JSONL for analysis.

## Basic Usage

Create a trace with spans to track an agent workflow:

```{r}
library(securetrace)

# Wrap your workflow in a trace
result <- with_trace("my-agent-run", {

  # Each operation gets its own span
  with_span("llm-call", type = "llm", {
    record_tokens(1500, 300, model = "claude-sonnet-4-5")
    "The answer is 42"
  })
})
```

## Token and Cost Tracking

securetrace includes built-in pricing for popular LLM models:

```{r}
# View known model pricing (per 1M tokens)
model_costs()

# Calculate cost for a specific call
calculate_cost("claude-opus-4-6", input_tokens = 5000, output_tokens = 1000)

# Register custom model pricing
add_model_cost("my-fine-tuned-model", input_price = 5, output_price = 20)
```

## Exporting Traces

Export completed traces to JSONL files or the console:
```{r}
# JSONL file export
exp <- jsonl_exporter("traces.jsonl")

with_trace("exported-run", exporter = exp, {
  with_span("work", type = "tool", {
    42
  })
})

# Console export for debugging
debug_exp <- console_exporter(verbose = TRUE)

# Combine multiple exporters
both <- multi_exporter(exp, debug_exp)

# Or set a default exporter for all traces
set_default_exporter(exp)
```

## Integration Helpers

securetrace provides helpers for common operations:

```{r}
# Wrap tool calls
result <- with_trace("tools", {
  trace_tool_call("calculator", function(x) x * 2, 21)
})

# Wrap guardrail checks
with_trace("guarded", {
  trace_guardrail("length-check", function(x) nchar(x) < 1000, user_input)
})
```

## Nested Spans

Spans can be nested to represent hierarchical operations:

```{r}
with_trace("complex-workflow", {
  with_span("planning", type = "llm", {
    record_tokens(2000, 500, model = "claude-sonnet-4-5")

    with_span("tool-use", type = "tool", {
      # Tool execution within the planning step
      42
    })
  })

  with_span("summarize", type = "llm", {
    record_tokens(1000, 200, model = "claude-haiku-4-5")
    "Summary complete"
  })
})
```

## Trace Summary

Get a formatted summary of any trace:

```{r}
tr <- Trace$new("manual-trace")
tr$start()
s <- Span$new("llm", type = "llm")
s$start()
s$set_tokens(input = 5000, output = 1000)
s$set_model("claude-opus-4-6")
s$end()
tr$add_span(s)
tr$end()

tr$summary()
# Trace: manual-trace (completed)
#   Duration: 0.01s
#   Spans: 1
#   Tokens: 5000 input, 1000 output
#   Cost: $0.150000
```
